Hierarchical Time Series Forecasting in Emergency Medical Services

### 2022-11-20 manuscript/01-introduction.qmd
final draft from bahman

However, despite a large number of studies dedicated to forecasting for EMS [@mingliterature2022; @gul2020exhaustive; @ibrahim2016modeling; @wargon2009systematic], the hierarchical data structure has been largely ignored, and the main focus has been on producing independent forecasts at a single level. Generating independent forecasts can result in a lack of consistency and coordination, and therefore leads to less effective planning and decision making. With hierarchical forecasting, plans at any level are based on coherent forecasts and therefore can be aligned. Implementing and sustaining improvements in EMS require alignments and coordination between different stakeholders, without which teams operate in isolation leading to conflicts, duplication work, rework, or work that runs counter to the overall goal to improve the quality of delivery service. Hierarchical forecasting framework can be used as a tool to improve coordination between teams across the care services at the national, sub-national, regional and local levels. The hierarchical forecasting approaches not only create consistent forecasts, but are usually also more accurate than the independent (base) forecasts [@hyndman2011optimal]. To our knowledge, there has been no previous research involving hierarchical and grouped forecasting in the entire field of forecasting for healthcare management.

In this paper, we address this gap by investigating the application of hierarchical forecasting approaches in the EMS using daily time series of verified incidents from 2015 to 2020 in a major ambulance service in Great Britain. The data has hierarchical and grouped structures, with hierarchies at the national, control (i.e. sub-national), health board (i.e. regional) levels, as well as groups by priority and nature of incidents. We produce consistent point forecasts and forecast distributions for all levels, which is critical for an effective planning and associated risk management. We compare the point and probabilistic forecast accuracy of the independent forecasts, bottom-up and optimal reconciliation approaches. We first generate independent/base forecasts using Exponential Smoothing State Space (ETS), Generalized Linear Model (GLM), Poisson regression, a simple empirical distribution and an ensemble method, followed by applying bottom-up and optimal reconciliation approaches. Forecast performance is assessed by the Mean Squared Scaled Error (MSSE) for point forecasts and Continuous Ranked Probability Scores (CRPS) for the probabilistic forecasts.
This paper complies with reproducibility principles [@stodden2013best; @boylan2015reproducibility], and can be applied to any healthcare service (e.g., emergency department, primary or social care) subject to the time series having a hierarchical and/or grouped structure.

>>>>

However, despite a large number of studies dedicated to forecasting for EMS [@mingliterature2022; @gul2020exhaustive; @ibrahim2016modeling; @wargon2009systematic], the hierarchical data structure has been largely ignored, and the main focus has been on producing independent (base) forecasts at a single level. Generating independent forecasts can result in a lack of consistency and coordination, and therefore leads to less effective planning and decision making. With hierarchical forecasting, plans at any level are based on coherent forecasts and therefore can be aligned. Implementing and sustaining improvements in EMS require alignments and coordination between different stakeholders, without which teams operate in isolation leading to conflicts, duplication work, rework, or work that runs counter to the overall goal to improve the quality of delivery service. Hierarchical forecasting framework can be used as a tool to improve coordination between teams across the care services at the national, sub-national, regional and local levels. The hierarchical forecasting approaches not only create consistent forecasts, but are usually also more accurate than the independent (base) forecasts [@hyndman2011optimal]. To our knowledge, there has been no previous research involving hierarchical and grouped forecasting in the entire field of forecasting for healthcare management.

In this paper, we address this gap by investigating the application of hierarchical forecasting approaches in the EMS using daily time series of attended incidents from 2015 to 2020 in a major ambulance service in Great Britain. The data has hierarchical and grouped structures, with hierarchies at the national, control (i.e. sub-national), health board (i.e. regional) levels, as well as groups by priority and nature of incidents. We produce consistent point forecasts and forecast distributions for all levels, which is critical for an effective planning and associated risk management. We compare the point and probabilistic forecast accuracy of the independent forecasts, bottom-up and optimal reconciliation approaches. We first generate independent/base forecasts using Exponential Smoothing State Space (ETS), Poisson regression using Generalized Linear Model (GLM) and tscount (TSGLM), a simple empirical distribution and an ensemble method, followed by applying bottom-up and optimal reconciliation approaches. Forecast performance is assessed by the Mean Absolute Scaled Error (MASE) and  Mean Squared Scaled Error (MSSE) for point forecasts and Continuous Ranked Probability Scores (CRPS) for the probabilistic forecasts.
This paper complies with reproducibility principles [@stodden2013best; @boylan2015reproducibility]. We provide the R codes for the proposed models and benchmarks. Therefore, they can be applied to any healthcare service (e.g., emergency department, primary or social care) subject to the time series having a hierarchical and/or grouped structure.

### 2022-11-20 manuscript/03-experiment.qmd
final draft from bahman

Given the presence of various significant patterns in the past attended incidents, we consider three different forecasting models to generate the base forecasts. Once the base forecasts are produced, hierarchical and grouped time series methods are used to reconcile them across all levels. We briefly discuss forecasting models in the following sections, and the hierarchical forecasting methods are discussed in @sec-htc.

@@

<!-- add a sentence to describe how the point forecast and forecast distribution is created for the ensemble -->

>>>>

Given the presence of various patterns in the past attended incidents, we consider three different forecasting models to generate the base forecasts. Once the base forecasts are produced, hierarchical and grouped time series methods are used to reconcile them across all levels. We briefly discuss forecasting models in the following sections, and the hierarchical forecasting methods are discussed in @sec-htc.

@@

### 2022-11-20 manuscript/04-htsgts.qmd
final draft from bahman

Let $\bm{b}_t$ be a vector of $n_b$ "bottom-level" time series at time $t$, and let $\bm{a}_t$ be a corresponding vector of $n_a = n-n_b$ aggregated time series, where
$$
  \bm{a}_t = \bm{A}\bm{b}_t,
$$
and $\bm{A}$ is the $n_a\times n_b$ "aggregation" matrix specifying how the bottom-level series $\bm{b}_t$ are to be aggregated to form $\bm{a}_t$. The full vector of time series is given by
$$
 \bm{y}_t = \begin{bmatrix}\bm{a}_t \\\bm{b}_t\end{bmatrix}.
$$
This leads to the $n\times n_b$ "summing" or "structural" matrix given by
$$
  \bm{S} = \begin{bmatrix}\bm{A} \\ \bm{I}_{n_b}\end{bmatrix}
$$
such that $\bm{y}_t = \bm{S}\bm{b}_t$.

>>>>

Let $\bm{b}_t$ be a vector of $n_b$ _bottom-level_ time series at time $t$, and let $\bm{a}_t$ be a corresponding vector of $n_a = n-n_b$ aggregated time series, where
$$
  \bm{a}_t = \bm{A}\bm{b}_t,
$$
and $\bm{A}$ is the $n_a\times n_b$ "aggregation" matrix specifying how the bottom-level series $\bm{b}_t$ are to be aggregated to form $\bm{a}_t$. The full vector of time series is given by
$$
 \bm{y}_t = \begin{bmatrix}\bm{a}_t \\\bm{b}_t\end{bmatrix}.
$$
This leads to the $n\times n_b$ "summing" or "structural" matrix given by
$$
  \bm{S} = \begin{bmatrix}\bm{A} \\ \bm{I}_{n_b}\end{bmatrix}
$$
such that $\bm{y}_t = \bm{S}\bm{b}_t$.

### 2022-11-20 manuscript/05-results.qmd
final draft from bahman

The overall forecasting performance is reported in @tbl-result, in which the average forecast accuracy over horizons 43--84 days (corresponding to the planning horizon) is presented per model, method and the hierarchical level. Reported forecast accuracy are averaged across all forecast horizons, rolling origins and series at each level. @tbl-result presents both point and probabilistic forecast accuracy at total, control area, health board and bottom level series. Point forecast performance are reported using MASE and MSSE in @tbl-result-1  and @tbl-result-2, respectively. Probabilistic forecast accuracy is reported using CRPS in @tbl-result-3. The bold entries in each table identify a combination of method and model that performs best for the corresponding level (i.e. each column), based on the smallest values of accuracy measures.

@tbl-result-1 and 2b show that forecast reconciliation (i.e. MinT) improves forecast accuracy at the higher levels of the hierarchy including total, control area and health board. However, it is does not result in accuracy improvement at the bottom level series, for which base forecasts are more accurate. This might be due to the noisy structure of time series at the bottom level, and possibly due to very different patterns in the aggregated series. It is also clear from @tbl-result-1 that the ensemble method improves forecast accuracy at total, control area and health board. However, this does not remain valid for bottom series where different individual methods perform best, depending on the accuracy measure.

@@

# Plot of average accuracy vs week for each method for Total
acc_summary <- accuracy |>
  filter(
    series %in% c("Total", "Control areas", "Health boards"),
    method == "mint", model != "qcomb"
  ) |>
  mutate(
    measure = factor(measure, levels = c("mase","msse","crps"), labels = c("MASE","MSSE","CRPS")),
    series = factor(series, levels = c("Total", "Control areas", "Health boards")),
    model = factor(model, levels = c("naiveecdf", "ets", "tscount", "iglm", "ensemble")),
    week = factor(trunc((h - 1) / 7) + 1)
  ) |>
  group_by(week, model, measure, series) |>
  summarise(accuracy = mean(accuracy), .groups = "drop")

>>>>

The overall forecasting performance is reported in @tbl-result, in which the average forecast accuracy over horizons 43--84 days (corresponding to the planning horizon) is presented per model, method and the hierarchical level. Reported forecast accuracy are averaged across all forecast horizons, rolling origins and series at each level. @tbl-result presents both point and probabilistic forecast accuracy at total, control area, health board and bottom level series. Point forecast performance are reported using MASE and MSSE in @tbl-result-1  and [-@tbl-result-2], respectively. Probabilistic forecast accuracy is reported using CRPS in @tbl-result-3. The bold entries in each table identify a combination of method and model that performs best for the corresponding level (i.e. each column), based on the smallest values of accuracy measures.

@tbl-result-1 and [-@tbl-result-2] show that forecast reconciliation (i.e. MinT) improves forecast accuracy at the higher levels of the hierarchy including total, control area and health board. However, it is does not result in accuracy improvement at the bottom level series, for which base forecasts are more accurate. This might be due to the noisy structure of time series at the bottom level, and possibly due to very different patterns in the aggregated series. It is also clear from @tbl-result-1 that the ensemble method improves forecast accuracy at total, control area and health board. However, this does not remain valid for bottom series where different individual methods perform best, depending on the accuracy measure.

@@

# Plot of average accuracy vs week for each method for Total
acc_summary <- accuracy |>
  filter(
    series %in% c("Total", "Control areas", "Health boards"),
    method == "mint", model != "qcomb"
  ) |>
  mutate(model = case_when(
    model == "naiveecdf" ~ "Naive",
    model == "ets" ~  "ETS",
    model == "tscount" ~ "TSGLM",
    model == "iglm" ~ "GLM",
    model == "ensemble" ~ "Ensemble"
  )) |> 
  mutate(
    measure = factor(measure, levels = c("mase","msse","crps"), labels = c("MASE","MSSE","CRPS")),
    series = factor(series, levels = c("Total", "Control areas", "Health boards")),
    model = factor(model, levels = c("Naive", "ETS", "TSGLM", "GLM", "Ensemble")),
    week = factor(trunc((h - 1) / 7) + 1)
  ) |>
  group_by(week, model, measure, series) |>
  summarise(accuracy = mean(accuracy), .groups = "drop")

### 2022-11-20 manuscript/06-conclusion.qmd
final draft from bahman

Forecasting problems at Emergency Medical Services often have inherent hierarchical and grouped structures. For example, looking at time series of arrival calls in a clinical desk service, verified incidents, or attended incidents in a country, they could be disaggregated by various attributes of interest. Total demand in the country could be disaggregated by region, then within each region by health board, within each health board, by station/hospital, and so on down to the post code area. Alternative structures may arise when attributes of interest are crossed rather than nested. For example, the total demand could be disaggregated by priority (e.g., Red, Amber, Green) or by nature of incidents. It is also natural to have a mixed structures, for example the total demand could be disaggregated by priority and also by health board.

@@

Further research could investigate the practical benefits of probabilistic hierarchical forecasting in EMS. Linking forecasts with its utilities (response time, resource utilization, etc) can offer an opportunity to maximize benefits through more holistic planning approach. While, we generated count attend incident for the base forecast using Poisson regression models, however the reconciles forecast is snot yet count. This could be also an avenue for further research.

>>>>

Forecasting problems at Emergency Medical Services often have inherent hierarchical and grouped structures. For example, looking at time series of arrival calls in a clinical desk service, Emergency Department admissions, verified incidents, or attended incidents in a country, they could be disaggregated by various attributes of interest. Total demand in the country could be disaggregated by region, then within each region by health board, within each health board, by station/hospital, and so on down to the post code area. Alternative structures may arise when attributes of interest are crossed rather than nested. For example, the total demand could be disaggregated by priority (e.g., Red, Amber, Green) or by nature of incidents. It is also natural to have a mixed structures, for example the total demand could be disaggregated by priority and also by health board.

@@

Further research could investigate the practical benefits of probabilistic hierarchical forecasting in EMS. Linking forecasts with its utilities (response time, resource utilization, etc) can offer an opportunity to maximize benefits through more holistic planning approach. While, we generated count attend incident for the base forecast using Poisson regression models, however the reconciles forecast is not yet count. This could be also an avenue for further research.

### 2022-11-20 manuscript/main.qmd
final draft from bahman

---
title: "Hierarchical Time Series Forecasting in Emergency Medical Services"
author:
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations:
        - id: carbs
          name: Cardiff University
          department: Cardiff Business School
          country: United Kingdom
          postal-code: CF10 3EU
    attributes:
        corresponding: true
  - name: Rob J. Hyndman
    email: Rob.Hyndman@monash.edu
    affiliations:
        - id: monash
          name: Monash University
          department: Department of Econometrics and Business Statistics
          country: Australia
          postal-code: VIC 3800
abstract: |
  Accurate forecasts of ambulance demand are crucial inputs when planning and deploying staff and fleet. Such demand forecasts are required at national, regional and sub-regional levels, and must take account of the nature of incidents and their priorities. These forecasts are often generated independently by different teams within the organization. As a result, forecasts at different levels may be inconsistent, resulting in conflicting decisions and a lack of coherent coordination in the service. To address this issue, we exploit the hierarchical and grouped structure of the demand time series, and apply forecast reconciliation methods to generate both point and probabilistic forecasts that are coherent and use all the available data at all levels of disaggregation. The methods are applied to daily incident data from an ambulance service in Great Britain, from October 2015 to July 2019, disaggregated by nature of incident, priority, managing health board, and control area. We use an ensemble of forecasting models, and show that the resulting forecasts are better than any individual forecasting model. We validate the forecasting approach using time-series cross-validation.
keywords:
  - healthcare
  - emergency services
  - forecast reconciliation
  - ambulance demand
  - regression
bibliography: bibliography.bib
execute:
  echo: false
  message: false
  warning: false
  cache: true
pdf-engine: pdflatex
format:
  elsevier-pdf:
    keep-tex: true
    include-in-header: preamble.tex
    journal:
      name: Annals of Emergency Medicine
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

>>>>

---
title: "Hierarchical Time Series Forecasting in Emergency Medical Services"
author:
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations:
        - id: carbs
          name: Cardiff University
          department: Cardiff Business School
          country: United Kingdom
          postal-code: CF10 3EU
    attributes:
        corresponding: true
  - name: Rob J. Hyndman
    email: Rob.Hyndman@monash.edu
    affiliations:
        - id: monash
          name: Monash University
          department: Department of Econometrics and Business Statistics
          country: Australia
          postal-code: VIC 3800
abstract: |
  Accurate forecasts of ambulance demand are crucial inputs when planning and deploying staff and fleet. Such demand forecasts are required at national, regional and sub-regional levels, and must take account of the nature of incidents and their priorities. These forecasts are often generated independently by different teams within the organization. As a result, forecasts at different levels may be inconsistent, resulting in conflicting decisions and a lack of coherent coordination in the service. To address this issue, we exploit the hierarchical and grouped structure of the demand time series, and apply forecast reconciliation methods to generate both point and probabilistic forecasts that are coherent and use all the available data at all levels of disaggregation. The methods are applied to daily incident data from an ambulance service in Great Britain, from October 2015 to July 2019, disaggregated by nature of incident, priority, managing health board, and control area. We use an ensemble of forecasting models, and show that the resulting forecasts are better than any individual forecasting model. We validate the forecasting approach using time-series cross-validation.
keywords:
  - healthcare
  - emergency services
  - forecast reconciliation
  - ambulance demand
  - regression
bibliography: bibliography.bib
execute:
  echo: false
  message: false
  warning: false
  cache: true
pdf-engine: pdflatex
format:
  elsevier-pdf:
    keep-tex: true
    include-in-header: 
      - text: \usepackage{bm}
      - file: preamble.tex
    journal:
      name: Annals of Emergency Medicine
      formatting: preprint
      model: 3p
      cite-style: authoryear
---
